Reducing RES Droughts through the integration of wind and PV

### 2024-11-29 Article/main.tex
CS edit complete

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\textwidth]{verification_wind_bar}
	\caption{Average annual number of wind drought events for EirGrid (black), Atlite (red), C3S-E G (blue), and C3S-E N (purple)}
	\label{fig:bar_number_events_verification_wind}
\end{figure}

@@

Figure \ref{fig:solar_verification_contour} shows that the three datasets have a similar tendency to overestimate the CF compared to EirGrid, especially for high CF values. The skill scores presented in Table~\ref{tab:pv_skill_scores} indicate that C3S-E G performs best overall, with the lowest RMSE and a high correlation coefficient, suggesting a closer match to observed data. Atlite shows a slightly lower correlation and a higher RMSE, indicating larger deviations. All models show a slight positive bias, with Atlite exhibiting the most overestimation. These results reflect the broader challenges in accurately modelling PV, where variability is hard to capture. Nonetheless, C3S-E G’s lower RMSE suggests it may be more reliable for PV CF modelling in this context.

@@

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\textwidth]{verification_pv_bar}
	\caption{Annual number of PV drought events for EirGrid (black), Atlite (red), C3S-E G (blue), and C3S-E N (purple) in 2023}
	\label{fig:bar_number_events_verification_pv}
\end{figure}

@@

The first part of the analysis examines the annual number of RES drought events across the three datasets.

For wind (Fig.~\ref{fig:boxplot_number_events}a), the number of events decreases as the duration range increases, with very few events lasting more than seven days. For PV (Fig.~\ref{fig:boxplot_number_events}b), the number of events also declines as the duration range extends from one to eight days, followed by a slight increase for longer durations. This increase is due to extended low-generation periods occurring from November to March, depending on the dataset. When comparing wind and PV results (Fig.~\ref{fig:boxplot_number_events}a \& Fig.~\ref{fig:boxplot_number_events}b), the median, first, and third quartiles for PV are consistently higher or equal to those for wind, across all duration ranges and datasets. This is due to the typically lower CF of PV compared to wind, especially in a region like Ireland where solar potential is limited. PV generation is also zero at night and constrained by the daily solar cycle, leading to a naturally higher frequency of RES droughts in PV compared to wind.

The bottom panels show the combination of wind and PV under the two capacity scenarios. In the existing RES installation scenario (Fig.~\ref{fig:boxplot_number_events}c), the identified RES droughts closely match those for wind alone, which is expected due to the dominance of installed wind capacity. In contrast, the balanced scenario (Fig.~\ref{fig:boxplot_number_events}d) shows a clear reduction in the number of drought events across all datasets and durations, with a decrease of the total number of events of 56\% for Atlite, 52\% for C3S-E G, and 50\% for C3S-E N. This reduction is attributed to the anti-correlation between wind and PV production throughout the year, a phenomenon well-documented in the literature \cite{kaspar2019drought}.

@@

The RES drought events identified over the 45-year period were used to calculate the return periods for different RES drought durations. A return period refers to the estimated average interval between occurrences of events with a specified duration or intensity. Fig.~\ref{fig:return_periods} illustrates the return periods for varying RES drought durations, highlighting how often different drought lengths are likely to occur across the datasets. This analysis provides insight into the frequency and likelihood of prolonged low-generation periods, which is crucial for evaluating the potential impact of RES droughts on energy reliability and security of supply.

@@

For PV (Fig.~\ref{fig:return_periods}b), Atlite behaves differently than the two C3S-E datasets. The Atlite results show a log-linear increase but reach higher values in general with the longest event lasting forty days. For C3S-E G and C3S-E N, the duration of RES droughts increases in a log-linear pattern for events lasting less than 16 days. Beyond this duration, there is a sharp rise in drought duration for events up to a one-year return period. This sudden increase reflects the impact of winter on PV generation in Ireland, as PV output often remains below the CF threshold for extended periods during winter months. The difference between Atlite and the C3S-E datasets arises from how each model handles PV output near the threshold of 0.1 CF. Atlite remains slightly above the threshold more frequently during these conditions, leading to shorter, more fragmented drought events. In contrast, C3S-E G and C3S-E N tend to fall below the threshold in similar conditions, resulting in longer continuous drought periods, especially during winter. This sensitivity to the threshold highlights how slight model differences can have substantial effects on drought duration estimates, particularly for PV in low-generation conditions.

@@

In this study, a threshold CF value of 0.1 was used to identify RES droughts. To assess how sensitive the results are to this threshold, the annual number of events was calculated across a range of CF threshold values (Fig.~\ref{fig:number_days_threshold}).

Across all four panels, the results reveal a general trend: as the threshold increases, so does the number of identified events. This increase, however, is not uniform across datasets and energy sources, but the overall pattern remains consistent. When the threshold increases, smaller events merge into larger, less frequent events. Depending on the dataset, the number of events reaches a peak around or after a threshold of 0.2, and then gradually decreases until only a single event encompasses the entire time series.

@@

This study highlighted the key differences between three RES models: Atlite, C3S-E G, and C3S-E N, in their ability to represent RES droughts. One of the most significant differences is how each dataset incorporates the specific locations of RES farms. Both Atlite and C3S-E G consider the locations of wind and PV farms, which should, in theory, provide a more accurate representation of RES generation. While this approach improves PV drought models, our analysis indicates that for wind, the Atlite dataset performs better overall, especially in its close alignment with EirGrid data for wind generation estimates. This finding suggests that although the inclusion of RES farm locations is beneficial, particularly for PV, the accuracy of the power curve model in Atlite may be the critical factor in producing realistic time series for low-generation events.

The results showed that the two C3S-E datasets (C3S-E G and C3S-E N) consistently yield similar outcomes, indicating that their methodological differences have minimal impact. In contrast, Atlite stands out for its closer alignment with EirGrid data, particularly in wind generation, as seen in the verification process (Section \ref{sec:wind_verification}). This distinction was also evident in the analysis, where Atlite reported higher return periods and a greater number of RES droughts, especially in scenarios with a balanced share of RES. One of the study's most critical findings is that the accuracy of RES drought modelling relies more on the precision of the wind power curve and PV panel models than on the specific locations of RES farms. Atlite’s superior performance highlights the importance of selecting accurate models for assessing RES drought risks. This careful model selection can better quantify risks, support effective planning, and avoid the potential underestimation of capacity needs, which is essential for ensuring energy security.

Looking at the balanced scenario, the analysis showed a significant improvement in the management of RES droughts due to the complementary nature of wind and PV generation. Wind and PV together perform better in terms of drought frequency and duration than either would individually, largely because of the seasonal anti-correlation between the two energy sources. This diversification reduces the seasonal impact on RES droughts, as PV generation peaks in the summer and wind generation is more consistent in winter. Ireland currently has a highly wind-dependent energy system, but with ambitious targets for PV installations in the coming years, the energy mix is expected to approach a balance between wind and PV capacity. This balanced approach offers a more stable and secure energy supply by mitigating RES drought risks. For policymakers, these findings underscore the importance of meeting these capacity targets to enhance energy security through diversification. Additionally, the choice of model for RES drought assessment becomes increasingly critical as more renewable capacity is integrated into the system.

In terms of future research, there are two main topics to explore. First, future climate data could be integrated with energy scenarios to better understand how climate change might affect RES droughts. Second, expanding the scope of the study to include Europe would provide a more comprehensive understanding of RES droughts in an interconnected energy grid. However, this expansion would require extensive verification across all EU countries, making it a more complex but highly relevant challenge.

>>>>

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\textwidth]{verification_wind_bar}
	\caption{Average annual number of wind drought events for EirGrid (black outline), Atlite (red), C3S-E G (blue), and C3S-E N (purple)}
	\label{fig:bar_number_events_verification_wind}
\end{figure}

@@

Figure \ref{fig:solar_verification_contour} shows that the three datasets have a similar tendency to overestimate the CF compared to EirGrid, especially for high CF values. The skill scores presented in Table~\ref{tab:pv_skill_scores} indicate that C3S-E~G performs best overall, with the lowest RMSE and a high correlation coefficient, suggesting a closer match to observed data. All models show a slight positive bias, with Atlite exhibiting a slightly lower correlation and higher RMSE. These results reflect the broader challenges in accurately modelling PV, where variability is hard to capture.

@@

\begin{figure}[!ht]
	\centering
	\includegraphics[width=\textwidth]{verification_pv_bar}
	\caption{Annual number of PV drought events for EirGrid (black outline), Atlite (red), C3S-E G (blue), and C3S-E N (purple) in 2023}
	\label{fig:bar_number_events_verification_pv}
\end{figure}

@@

The first part of the analysis examines the annual number of RES drought events across the three datasets. For wind (Fig.~\ref{fig:boxplot_number_events}a), the number of events decreases as the duration range increases, with very few events lasting more than seven days. For PV (Fig.~\ref{fig:boxplot_number_events}b), the number of events also declines as the duration range extends from one to eight days, followed by a slight increase for longer durations. This increase is due to extended low-generation periods occurring from November to March, depending on the dataset. When comparing wind and PV results (Fig.~\ref{fig:boxplot_number_events}a \& Fig.~\ref{fig:boxplot_number_events}b), the median, first, and third quartiles for PV are consistently higher or equal to those for wind, across all duration ranges and datasets. This is due to the typically lower CF of PV compared to wind, especially in a region like Ireland where solar potential is limited. PV generation is also zero at night and constrained by the daily solar cycle, leading to a naturally higher frequency of RES droughts in PV compared to wind.

The bottom panels show the combination of wind and PV under the two capacity scenarios. In the existing RES installation scenario (Fig.~\ref{fig:boxplot_number_events}c), the identified RES droughts closely match those for wind alone, which is expected due to the dominance of installed wind capacity. In contrast, the balanced scenario (Fig.~\ref{fig:boxplot_number_events}d) shows a clear reduction in the number of drought events across all datasets and durations, with a decrease of the total number of events of 56\% for Atlite, 52\% for C3S-E G, and 50\% for C3S-E N. This reduction is attributed to the anti-correlation between wind and PV production.

@@

The RES drought events identified over the 45-year period were used to calculate the return periods for different RES drought durations. A return period refers to the estimated expected interval between occurrences of events with a specified duration or intensity. Fig.~\ref{fig:return_periods} illustrates the return periods for varying RES drought durations, highlighting how often different drought lengths are likely to occur across the datasets. This analysis provides insight into the frequency and likelihood of prolonged low-generation periods, which is crucial for evaluating the potential impact of RES droughts on energy reliability and security of supply.

@@

For PV (Fig.~\ref{fig:return_periods}b), Atlite behaves differently than the two C3S-E datasets. The Atlite results show a log-linear increase but reach higher values in general with the longest event lasting forty days. For C3S-E G and C3S-E N, the duration of RES droughts increases in a log-linear pattern for events lasting less than 16 days. Beyond this duration, there is a sharp rise in drought duration for events up to a one-year return period. This sudden increase reflects the impact of winter on PV generation in Ireland, as PV output often remains below the CF threshold for extended periods during winter months. The difference between Atlite and the C3S-E results arises from differences in the datasets near the threshold of 0.1 CF. Atlite remains slightly above the threshold more frequently during these conditions, leading to shorter, more fragmented drought events. In contrast, C3S-E G and C3S-E N tend to fall below the threshold in similar conditions, resulting in longer continuous drought periods, especially during winter. This sensitivity to the threshold highlights how slight model differences can have substantial effects on drought duration estimates, particularly for PV in low-generation conditions.

@@

In this study, a threshold CF value of 0.1 was used to identify RES droughts. To assess how sensitive the results are to this threshold, the annual number of RES drought events was calculated for a range of CF threshold values (Fig.~\ref{fig:number_days_threshold}). The results reveal a general trend: as the threshold increases, so too does the number of identified events. This increase is not uniform across datasets and energy sources, but the overall pattern remains consistent. When the threshold increases, smaller events merge into larger, less frequent events. Depending on the dataset, the number of events reaches a peak around or after a threshold of 0.2, and then gradually decreases until only a single event encompasses the entire time series.

@@

This study has investigated the ability of three RES models to represent RES droughts: Atlite, C3S-E G, and C3S-E N. One of the most evident differences is how each dataset incorporates the specific locations of RES farms. Both Atlite and C3S-E G consider the locations of wind and PV farms, which should, in theory, provide a more accurate representation of RES generation. While this approach slightly improves PV models, our analysis indicates that for wind, the Atlite dataset performs better overall, especially in its close alignment with EirGrid data for wind generation estimates. This finding suggests that, although the inclusion of RES farm locations is beneficial, the accuracy of the RES model is more strongly influenced by underlying model assumptions, such as selecting an appropriate wind power curve.

Atlite shows the best alignment with EirGrid data for wind generation. Differences between the models are smaller for PV, with C3S-G performing marginally better than the other two. The results show that the two C3S-E datasets (C3S-E G and C3S-E N) consistently yield similar outcomes, indicating that their methodological differences have minimal impact. This distinction was also evident in the analysis, where Atlite reported higher return periods and a greater number of RES droughts, especially in scenarios with a balanced share of RES. Again, the results from RES drought modelling relies more on the precision of the wind power curve and PV panel models than on the specific locations of RES farms. Atlite’s superior performance highlights the importance of selecting validated models for assessing RES drought risks. This careful model selection can better quantify risks, support effective planning, and avoid the potential underestimation of capacity needs, which is essential for ensuring energy security.

Looking at the balanced scenario, the analysis showed a significant improvement in the management of RES droughts due to the complementary nature of wind and PV generation. Wind and PV together perform better in terms of reducing drought frequency and duration than either would individually, largely because of the seasonal anti-correlation between the two energy sources. This diversification reduces the seasonal impact on RES droughts, as PV generation peaks in the summer and wind generation is more consistent in winter. Ireland currently has a highly wind-dependent energy system, but with ambitious targets for PV installations in the coming years, the energy mix is expected to approach a balance between wind and PV capacity. This balanced approach offers a more stable and secure energy supply by mitigating RES drought risks. For policymakers, these findings underscore the importance of meeting these capacity targets to enhance energy security through diversification. Additionally, the choice of model for RES drought assessment becomes increasingly critical as more renewable capacity is integrated into the system.

Future work is planned to extend the current analysis. First, climate projection data will be integrated with energy scenarios to better understand how climate change might affect RES droughts. Second, expanding the geographic domain of the study to include the EU would provide a more comprehensive understanding of RES droughts in an interconnected energy grid. This would require extensive verification across all EU countries, making it a more complex but highly relevant challenge.
